### 2018-01-10 submission/manuscript.Rmd
correct one small missing statement in methods
fix grammar errors in intro updates

Although there has been an intense focus on microbiota based biomarker discovery for CRC, the candidate genera seem to be numerous. Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Yet, even in some of these studies mouth-associated genera are far from the only microbes identified to be associated with CRC. These other genera include, but are not necessarily limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. In fact, *Escherichia/Shigella* also has good *in vivo* evidence that it could be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012]. Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential additional markers of CRC with *Bacteroides fragilis* having good mechanistic studies into how this may occur [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. Still, other studies have identified genera like *Roseburia* as increased in CRC while others have reported it to be decreased or have found no difference in any genera at all [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].

Most of these studies have focused on carcinoma but the adenoma results are not any clearer at identifying candidate genera. Some groups have focused on more broad scale community metrics and have found that metrics such as richness had decreased in adenoma versus controls while others have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017]. A recent meta-analysis confirmed the associations of mouth-associated genera and *Akkermansia muciniphila* in carcinoma [@shah_leveraging_2017]. However, it only analyzed a total of 509 stool samples and with some recent studies, that were not included, matching that sample size alone it is hard to know how extrapolatable these findings truly are. This meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) that need to be investigated further since a number of these genera have been found to be enriched in controls and not lesion (adenoma or carcinoma) [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015].

@@

***Study Analysis Overview:***
$\alpha$-diversity was first assessed for differences between controls, adenoma, and carcinoma using both linear mixed-effect models and RR. $\beta$-diversity was then assessed for each individual study for differences between control-adenoma and control-carcinoma. Next, all common genera were assessed for differences in RR for adenoma and carcinoma. We then built Random Forest models based on all or a selected model, based on the top 5 increased and top 5 decreased RR based on P-value, and these models were trained on one study then tested on the remaining studies, for every study. A similar approach was then applied at the OTU level with the exception that a 10-fold CV over 100 different models, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs. For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a variable in the top 5 increased and top 5 decreased RR based on P-value. Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power. For comparisons in which normal versus adenoma were made the carcinoma samples were excluded from each respective study. Similarly, for comparisons in which normal versus carcinoma were made the adenoma samples were excluded from each respective study. The data was split between feces and tissue samples. Within the tissue groups the data was further divided between samples from the same individual (matched) and those from different individuals (unmatched). Where applicable for each study, predictions for adenoma and carcinoma were then tested for feces, matched tissue, and unmatched tissue.

>>>>

Although there has been an intense focus on microbiota based biomarker discovery for CRC, the candidate genera seem to be numerous. Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Yet, even in some of these studies mouth-associated genera are far from the only microbes identified to be associated with CRC. These other genera include, but are not necessarily limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. In fact, *Escherichia/Shigella* and *Streptococcus* also have good *in vivo* evidence that individual species within the genus can be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017]. Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential additional markers of CRC with *Bacteroides fragilis* having good mechanistic studies into how this may occur [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. Still, other studies have identified genera like *Roseburia* as increased in CRC while others have reported it to be decreased or have found no difference in any genera at all [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].

Most of these studies have focused on carcinoma but the adenoma observations are not any clearer at identifying candidate genera. Some groups have focused on more broad scale community metrics and have found that metrics such as richness are decreased in adenoma versus controls while others have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017]. A recent meta-analysis confirmed the correlations of certain mouth-associated genera and *Akkermansia muciniphila* with carcinoma [@shah_leveraging_2017]. However, it only analyzed a total of 509 stool samples and with some recent studies, that were not included, matching this sample size alone, it is hard to know how extrapolatable these findings truly are. That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) that need to be investigated further, since a number of these genera have been found to be enriched in controls and not lesion (adenoma or carcinoma) [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015].

@@

***Study Analysis Overview:***
$\alpha$-diversity was first assessed for differences between controls, adenoma, and carcinoma using both linear mixed-effect models and RR. $\beta$-diversity was then assessed for each individual study for differences between control-adenoma and control-carcinoma. Next, all common genera were assessed for differences in RR for adenoma and carcinoma and ranked based on P-value. We then built Random Forest models based on all or a selected model, based on the top 5 increased and top 5 decreased RR based on P-value, and these models were trained on one study then tested on the remaining studies, for every study. A similar approach was then applied at the OTU level with the exception that a 10-fold CV over 100 different models, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs. For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a variable in the top 5 increased and top 5 decreased RR based on P-value. Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power. For comparisons in which normal versus adenoma were made the carcinoma samples were excluded from each respective study. Similarly, for comparisons in which normal versus carcinoma were made the adenoma samples were excluded from each respective study. The data was split between feces and tissue samples. Within the tissue groups the data was further divided between samples from the same individual (matched) and those from different individuals (unmatched). Where applicable for each study, predictions for adenoma and carcinoma were then tested for feces, matched tissue, and unmatched tissue.

### 2018-01-10 submission/manuscript.tex
correct one small missing statement in methods
fix grammar errors in intro updates

Although there has been an intense focus on microbiota based biomarker
discovery for CRC, the candidate genera seem to be numerous. Some
studies point towards mouth-associated genera such as
\emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas}, and
\emph{Porphyromonas} as key enriched genera {[}6,12--18{]}. Yet, even in
some of these studies mouth-associated genera are far from the only
microbes identified to be associated with CRC. These other genera
include, but are not necessarily limited to, \emph{Providencia},
\emph{Mogibacterium}, \emph{Enterococcus}, \emph{Escherichia/Shigella},
\emph{Klebsiella}, and \emph{Streptococcus} {[}14--16{]}. In fact,
\emph{Escherichia/Shigella} also has good \emph{in vivo} evidence that
it could be important in the pathogenesis of CRC {[}5,19{]}. Other
studies have also identified \emph{Akkermansia muciniphila} and
\emph{Bacteroides fragilis} as potential additional markers of CRC with
\emph{Bacteroides fragilis} having good mechanistic studies into how
this may occur {[}15,20,21{]}. Still, other studies have identified
genera like \emph{Roseburia} as increased in CRC while others have
reported it to be decreased or have found no difference in any genera at
all {[}15,18,22,23{]}.

Most of these studies have focused on carcinoma but the adenoma results
are not any clearer at identifying candidate genera. Some groups have
focused on more broad scale community metrics and have found that
metrics such as richness had decreased in adenoma versus controls while
others have identified \emph{Lactococcus}, \emph{Pseudomonas},
\emph{Acidovorax}, \emph{Cloacibacterium}, \emph{Helicobacter},
\emph{Lactobacillus}, \emph{Bilophila}, \emph{Desulfovibrio}, and
\emph{Mogibacterium} to be increased in adenoma {[}24--26{]}. A recent
meta-analysis confirmed the associations of mouth-associated genera and
\emph{Akkermansia muciniphila} in carcinoma {[}11{]}. However, it only
analyzed a total of 509 stool samples and with some recent studies, that
were not included, matching that sample size alone it is hard to know
how extrapolatable these findings truly are. This meta-analysis also
added more potential microbial associations to both carcinoma
(\emph{Pantoea agglomerans} \emph{Ruminococcus}, \emph{Lactobacillus})
and adenoma (\emph{Prevotella}, \emph{Methanosphaera},
\emph{Succinovibrio}, \emph{Haemophilus parainfluenzae},
\emph{Ruminococcus}, \emph{Lactobacillus}) that need to be investigated
further since a number of these genera have been found to be enriched in
controls and not lesion (adenoma or carcinoma) {[}13,16,17{]}.

Targeting the identification of biomarkers within stool seems logical
since it offers an easy and cost-effective way to stratify risk and the
current gold standard for diagnosis, a colonoscopy, can be
time-consuming and is not risk free. Although stool represents an easy
and less invasive way to assess risk, it is not clear how reflective
this sample is to the community on the adenoma or carcinoma. Some
studies have tried to assess this in health and disease but are limited
by their sample size {[}18,27{]}. Sampling the tissue directly may
provide clearer answers but is not without limitations. Due to bowel
prep the communities left for sampling may not be reflective of the
resident microbiota, but rather a collection of what is able to keep
adhered to the mucosa. Additionally, these samples contain more host
DNA, potentially limiting the types of analysis that can be done. It is
well known that low biomass samples can be very difficult to work with
and results can end up being study dependent due to the randomness of
contamination {[}28{]}. Due to these many differences that could arise
between stool and tissue, one question our meta-analysis aims to answer
is whether there are consistent patterns that emerge across studies
regardless of whether they used stool or tissue samples.

@@

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12--18,20,22--26,29{]} {[}Table 1 \& 2{]}, we expand both the breadth
and scope of the previous meta-analysis to investigate whether the
bacterial community is an important risk factor or if select members can
recapitulate the same information for both adenoma and carcinoma. To
accomplish this we first assessed whether bacterial diversity changes
throughout disease (control to adenoma to carcinoma) and if it results
in an increased relative risk (RR) for adenoma or carcinoma. We then
assessed what genera, if any, increase or decrease the RR of adenoma or
carcinoma. Next, using Random Forest models, we analyzed whether the
full community or only the top 5 increased and top 5 decreased RR genera
combined resulted in better model classification, based on the area
under the curve (AUC). Our results suggest that the community changes as
disease severity worsens and that this community is important for
disease classification. Finally, we also examined at what effect and
sample size the studies that were used were powered for and the sample
size needed to get to an 80\% power. Our power analysis suggests that
each individual study used was underpowered for detecting effect size
differences of 10\% or below between the case and control groups.

@@

One important caveat to this study is that even though specific genera
associated with species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean these specific species are not
important in human CRC {[}21,30{]}. Since we are limited in our
aggregation of the data, across study, to the genus level, it is not
possible to clearly delineate what specific species are contributing to
overall disease progression. However, \emph{Streptococcus gallolyticus}
subsp. \emph{gallolyticus} is a mouth-associated micobe and our study
clearly shows that regardless of whether the sample is tissue or stool,
mouth-associated genera are commonly associated with an increased RR of
both adenoma and carcinoma.

@@

\textbf{\emph{Obtaining Data Sets:}} The studies used for this
meta-analysis were identified through the review articles written by
Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}31,32{]} and
additional studies not mentioned in the reviews were obtained based on
the authors' knowledge of the literature. Studies that used tissue or
feces as their sample source for 454 or Illumina 16S rRNA gene
sequencing analysis were included and only data sets that had sequences
available for analysis were included. Some studies were excluded because
they did not have publicly available sequences or did not have metadata
in which the authors were able to share and were excluded. After these
filtering steps, the following studies remained: Ahn, \emph{et al.}
{[}12{]}, Baxter, \emph{et al.} {[}13{]}, Brim, \emph{et al.} {[}29{]},
Burns, \emph{et al.} {[}16{]}, Chen, \emph{et al.} {[}14{]}, Dejea,
\emph{et al.} {[}23{]}, Flemer, \emph{et al.} {[}18{]}, Geng, \emph{et
al.} {[}22{]}, Hale, \emph{et al.} {[}26{]}, Kostic, \emph{et al.}
{[}33{]}, Lu, \emph{et al.} {[}25{]}, Sanapareddy, \emph{et al.}
{[}24{]}, Wang, \emph{et al.} {[}15{]}, Weir, \emph{et al.} {[}20{]},
and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}34{]} study was not
included because the 90 individuals analyzed within the study are
contained within the larger Baxter study {[}13{]}. After sequence
processing all the case samples for the Kostic study had 100 or less
sequences remaining and was not used which left a total of 14 studies
that analysis could be completed on.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
(n = 1, {[}12{]}) and directly from the authors (n = 4,
{[}18,20,24,26{]}). Each study was processed using the mothur (v1.39.3)
software program {[}35{]} and quality filtering utilized the default
methods, used in mothur, for either 454 or Illumina based sequencing. If
it was not possible to use the defaults, the stated quality cut-offs,
from the study itself, were used instead. Chimeras were identified and
removed using VSEARCH {[}36{]} before \emph{de novo} OTU clustering at
97\% similarity was completed using the OptiClust algorithm {[}37{]}.

@@

\textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
assessed for differences between controls, adenoma, and carcinoma using
both linear mixed-effect models and RR. \(\beta\)-diversity was then
assessed for each individual study for differences between
control-adenoma and control-carcinoma. Next, all common genera were
assessed for differences in RR for adenoma and carcinoma. We then built
Random Forest models based on all or a selected model, based on the top
5 increased and top 5 decreased RR based on P-value, and these models
were trained on one study then tested on the remaining studies, for
every study. A similar approach was then applied at the OTU level with
the exception that a 10-fold CV over 100 different models, based on
random 80/20 splitting of the data, was used to generate a range of
expected AUCs. For these OTU-based models, the selected model included
all OTUs that had a taxonomic classification to a variable in the top 5
increased and top 5 decreased RR based on P-value. Finally, the power of
each study was assessed for an effect size ranging from 1\% to 30\% and
an estimated sample size, for these effect sizes, was generated based on
80\% power. For comparisons in which normal versus adenoma were made the
carcinoma samples were excluded from each respective study. Similarly,
for comparisons in which normal versus carcinoma were made the adenoma
samples were excluded from each respective study. The data was split
between feces and tissue samples. Within the tissue groups the data was
further divided between samples from the same individual (matched) and
those from different individuals (unmatched). Where applicable for each
study, predictions for adenoma and carcinoma were then tested for feces,
matched tissue, and unmatched tissue.

@@

\hypertarget{ref-weir_stool_2013}{}
20. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
Stool microbiome and metabolome differences between colorectal cancer
patients and healthy adults. PloS One. 2013;8:e70803.

\hypertarget{ref-bfrag_Boleij_2014}{}
21. Boleij A, Hechenbleikner EM, Goodwin AC, Badani R, Stein EM, Lazarev
MG, et al. The bacteroides fragilis toxin gene is prevalent in the colon
mucosa of colorectal cancer patients. Clinical Infectious Diseases
{[}Internet{]}. Oxford University Press (OUP); 2014;60:208--15.
Available from: \url{https://doi.org/10.1093/cid/ciu787}

\hypertarget{ref-geng_diversified_2013}{}
22. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.

\hypertarget{ref-dejea_microbiota_2014}{}
23. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
Rossetti BJ, et al. Microbiota organization is a distinct feature of
proximal colorectal cancers. Proceedings of the National Academy of
Sciences of the United States of America. 2014;111:18321--6.

\hypertarget{ref-sanapareddy_increased_2012}{}
24. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
et al. Increased rectal microbial richness is associated with the
presence of colorectal adenomas in humans. The ISME journal.
2012;6:1858--68.

\hypertarget{ref-lu_mucosal_2016}{}
25. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
adherent bacterial dysbiosis in patients with colorectal adenomas.
Scientific Reports. 2016;6:26337.

\hypertarget{ref-hale_shifts_2017}{}
26. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
Cancer Epidemiology, Biomarkers \& Prevention: A Publication of the
American Association for Cancer Research, Cosponsored by the American
Society of Preventive Oncology. 2017;26:85--94.

\hypertarget{ref-Flynn_preprint_2017}{}
27. Flynn KJ, Ruffin MT, Turgeon DK, Schloss PD. Spatial variation of
the native colon microbiota in healthy adults. Cold Spring Harbor
Laboratory; 2017; Available from: \url{https://doi.org/10.1101/189886}

\hypertarget{ref-Salter_contamination_2014}{}
28. Salter SJ, Cox MJ, Turek EM, Calus ST, Cookson WO, Moffatt MF, et
al. Reagent and laboratory contamination can critically impact
sequence-based microbiome analyses. BMC Biology {[}Internet{]}. Springer
Nature; 2014;12. Available from:
\url{https://doi.org/10.1186/s12915-014-0087-z}

\hypertarget{ref-brim_microbiome_2013}{}
29. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
al. Microbiome analysis of stool samples from African Americans with
colon polyps. PloS One. 2013;8:e81352.

\hypertarget{ref-strep_Aymeric_2017}{}
30. Aymeric L, Donnadieu F, Mulet C, Merle L du, Nigro G, Saffarian A,
et al. Colorectal cancer specific conditions promoteStreptococcus
gallolyticusgut colonization. Proceedings of the National Academy of
Sciences {[}Internet{]}. Proceedings of the National Academy of
Sciences; 2017;115:E283--91. Available from:
\url{https://doi.org/10.1073/pnas.1715112115}

>>>>

Although there has been an intense focus on microbiota based biomarker
discovery for CRC, the candidate genera seem to be numerous. Some
studies point towards mouth-associated genera such as
\emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas}, and
\emph{Porphyromonas} as key enriched genera {[}6,12--18{]}. Yet, even in
some of these studies mouth-associated genera are far from the only
microbes identified to be associated with CRC. These other genera
include, but are not necessarily limited to, \emph{Providencia},
\emph{Mogibacterium}, \emph{Enterococcus}, \emph{Escherichia/Shigella},
\emph{Klebsiella}, and \emph{Streptococcus} {[}14--16{]}. In fact,
\emph{Escherichia/Shigella} and \emph{Streptococcus} also have good
\emph{in vivo} evidence that individual species within the genus can be
important in the pathogenesis of CRC {[}5,19,20{]}. Other studies have
also identified \emph{Akkermansia muciniphila} and \emph{Bacteroides
fragilis} as potential additional markers of CRC with \emph{Bacteroides
fragilis} having good mechanistic studies into how this may occur
{[}15,21,22{]}. Still, other studies have identified genera like
\emph{Roseburia} as increased in CRC while others have reported it to be
decreased or have found no difference in any genera at all
{[}15,18,23,24{]}.

Most of these studies have focused on carcinoma but the adenoma
observations are not any clearer at identifying candidate genera. Some
groups have focused on more broad scale community metrics and have found
that metrics such as richness are decreased in adenoma versus controls
while others have identified \emph{Lactococcus}, \emph{Pseudomonas},
\emph{Acidovorax}, \emph{Cloacibacterium}, \emph{Helicobacter},
\emph{Lactobacillus}, \emph{Bilophila}, \emph{Desulfovibrio}, and
\emph{Mogibacterium} to be increased in adenoma {[}25--27{]}. A recent
meta-analysis confirmed the correlations of certain mouth-associated
genera and \emph{Akkermansia muciniphila} with carcinoma {[}11{]}.
However, it only analyzed a total of 509 stool samples and with some
recent studies, that were not included, matching this sample size alone,
it is hard to know how extrapolatable these findings truly are. That
particular meta-analysis also added more potential microbial
associations to both carcinoma (\emph{Pantoea agglomerans}
\emph{Ruminococcus}, \emph{Lactobacillus}) and adenoma
(\emph{Prevotella}, \emph{Methanosphaera}, \emph{Succinovibrio},
\emph{Haemophilus parainfluenzae}, \emph{Ruminococcus},
\emph{Lactobacillus}) that need to be investigated further, since a
number of these genera have been found to be enriched in controls and
not lesion (adenoma or carcinoma) {[}13,16,17{]}.

Targeting the identification of biomarkers within stool seems logical
since it offers an easy and cost-effective way to stratify risk and the
current gold standard for diagnosis, a colonoscopy, can be
time-consuming and is not risk free. Although stool represents an easy
and less invasive way to assess risk, it is not clear how reflective
this sample is to the community on the adenoma or carcinoma. Some
studies have tried to assess this in health and disease but are limited
by their sample size {[}18,28{]}. Sampling the tissue directly may
provide clearer answers but is not without limitations. Due to bowel
prep the communities left for sampling may not be reflective of the
resident microbiota, but rather a collection of what is able to keep
adhered to the mucosa. Additionally, these samples contain more host
DNA, potentially limiting the types of analysis that can be done. It is
well known that low biomass samples can be very difficult to work with
and results can end up being study dependent due to the randomness of
contamination {[}29{]}. Due to these many differences that could arise
between stool and tissue, one question our meta-analysis aims to answer
is whether there are consistent patterns that emerge across studies
regardless of whether they used stool or tissue samples.

@@

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12--18,21,23--27,30{]} {[}Table 1 \& 2{]}, we expand both the breadth
and scope of the previous meta-analysis to investigate whether the
bacterial community is an important risk factor or if select members can
recapitulate the same information for both adenoma and carcinoma. To
accomplish this we first assessed whether bacterial diversity changes
throughout disease (control to adenoma to carcinoma) and if it results
in an increased relative risk (RR) for adenoma or carcinoma. We then
assessed what genera, if any, increase or decrease the RR of adenoma or
carcinoma. Next, using Random Forest models, we analyzed whether the
full community or only the top 5 increased and top 5 decreased RR genera
combined resulted in better model classification, based on the area
under the curve (AUC). Our results suggest that the community changes as
disease severity worsens and that this community is important for
disease classification. Finally, we also examined at what effect and
sample size the studies that were used were powered for and the sample
size needed to get to an 80\% power. Our power analysis suggests that
each individual study used was underpowered for detecting effect size
differences of 10\% or below between the case and control groups.

@@

One important caveat to this study is that even though specific genera
associated with species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean these specific species are not
important in human CRC {[}20,22{]}. Since we are limited in our
aggregation of the data, across study, to the genus level, it is not
possible to clearly delineate what specific species are contributing to
overall disease progression. However, \emph{Streptococcus gallolyticus}
subsp. \emph{gallolyticus} is a mouth-associated micobe and our study
clearly shows that regardless of whether the sample is tissue or stool,
mouth-associated genera are commonly associated with an increased RR of
both adenoma and carcinoma.

@@

\textbf{\emph{Obtaining Data Sets:}} The studies used for this
meta-analysis were identified through the review articles written by
Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}31,32{]} and
additional studies not mentioned in the reviews were obtained based on
the authors' knowledge of the literature. Studies that used tissue or
feces as their sample source for 454 or Illumina 16S rRNA gene
sequencing analysis were included and only data sets that had sequences
available for analysis were included. Some studies were excluded because
they did not have publicly available sequences or did not have metadata
in which the authors were able to share and were excluded. After these
filtering steps, the following studies remained: Ahn, \emph{et al.}
{[}12{]}, Baxter, \emph{et al.} {[}13{]}, Brim, \emph{et al.} {[}30{]},
Burns, \emph{et al.} {[}16{]}, Chen, \emph{et al.} {[}14{]}, Dejea,
\emph{et al.} {[}24{]}, Flemer, \emph{et al.} {[}18{]}, Geng, \emph{et
al.} {[}23{]}, Hale, \emph{et al.} {[}27{]}, Kostic, \emph{et al.}
{[}33{]}, Lu, \emph{et al.} {[}26{]}, Sanapareddy, \emph{et al.}
{[}25{]}, Wang, \emph{et al.} {[}15{]}, Weir, \emph{et al.} {[}21{]},
and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}34{]} study was not
included because the 90 individuals analyzed within the study are
contained within the larger Baxter study {[}13{]}. After sequence
processing all the case samples for the Kostic study had 100 or less
sequences remaining and was not used which left a total of 14 studies
that analysis could be completed on.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
(n = 1, {[}12{]}) and directly from the authors (n = 4,
{[}18,21,25,27{]}). Each study was processed using the mothur (v1.39.3)
software program {[}35{]} and quality filtering utilized the default
methods, used in mothur, for either 454 or Illumina based sequencing. If
it was not possible to use the defaults, the stated quality cut-offs,
from the study itself, were used instead. Chimeras were identified and
removed using VSEARCH {[}36{]} before \emph{de novo} OTU clustering at
97\% similarity was completed using the OptiClust algorithm {[}37{]}.

@@

\textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
assessed for differences between controls, adenoma, and carcinoma using
both linear mixed-effect models and RR. \(\beta\)-diversity was then
assessed for each individual study for differences between
control-adenoma and control-carcinoma. Next, all common genera were
assessed for differences in RR for adenoma and carcinoma and ranked
based on P-value. We then built Random Forest models based on all or a
selected model, based on the top 5 increased and top 5 decreased RR
based on P-value, and these models were trained on one study then tested
on the remaining studies, for every study. A similar approach was then
applied at the OTU level with the exception that a 10-fold CV over 100
different models, based on random 80/20 splitting of the data, was used
to generate a range of expected AUCs. For these OTU-based models, the
selected model included all OTUs that had a taxonomic classification to
a variable in the top 5 increased and top 5 decreased RR based on
P-value. Finally, the power of each study was assessed for an effect
size ranging from 1\% to 30\% and an estimated sample size, for these
effect sizes, was generated based on 80\% power. For comparisons in
which normal versus adenoma were made the carcinoma samples were
excluded from each respective study. Similarly, for comparisons in which
normal versus carcinoma were made the adenoma samples were excluded from
each respective study. The data was split between feces and tissue
samples. Within the tissue groups the data was further divided between
samples from the same individual (matched) and those from different
individuals (unmatched). Where applicable for each study, predictions
for adenoma and carcinoma were then tested for feces, matched tissue,
and unmatched tissue.

@@

\hypertarget{ref-strep_Aymeric_2017}{}
20. Aymeric L, Donnadieu F, Mulet C, Merle L du, Nigro G, Saffarian A,
et al. Colorectal cancer specific conditions promoteStreptococcus
gallolyticusgut colonization. Proceedings of the National Academy of
Sciences {[}Internet{]}. Proceedings of the National Academy of
Sciences; 2017;115:E283--91. Available from:
\url{https://doi.org/10.1073/pnas.1715112115}

\hypertarget{ref-weir_stool_2013}{}
21. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
Stool microbiome and metabolome differences between colorectal cancer
patients and healthy adults. PloS One. 2013;8:e70803.

\hypertarget{ref-bfrag_Boleij_2014}{}
22. Boleij A, Hechenbleikner EM, Goodwin AC, Badani R, Stein EM, Lazarev
MG, et al. The bacteroides fragilis toxin gene is prevalent in the colon
mucosa of colorectal cancer patients. Clinical Infectious Diseases
{[}Internet{]}. Oxford University Press (OUP); 2014;60:208--15.
Available from: \url{https://doi.org/10.1093/cid/ciu787}

\hypertarget{ref-geng_diversified_2013}{}
23. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.

\hypertarget{ref-dejea_microbiota_2014}{}
24. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
Rossetti BJ, et al. Microbiota organization is a distinct feature of
proximal colorectal cancers. Proceedings of the National Academy of
Sciences of the United States of America. 2014;111:18321--6.

\hypertarget{ref-sanapareddy_increased_2012}{}
25. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
et al. Increased rectal microbial richness is associated with the
presence of colorectal adenomas in humans. The ISME journal.
2012;6:1858--68.

\hypertarget{ref-lu_mucosal_2016}{}
26. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
adherent bacterial dysbiosis in patients with colorectal adenomas.
Scientific Reports. 2016;6:26337.

\hypertarget{ref-hale_shifts_2017}{}
27. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
Cancer Epidemiology, Biomarkers \& Prevention: A Publication of the
American Association for Cancer Research, Cosponsored by the American
Society of Preventive Oncology. 2017;26:85--94.

\hypertarget{ref-Flynn_preprint_2017}{}
28. Flynn KJ, Ruffin MT, Turgeon DK, Schloss PD. Spatial variation of
the native colon microbiota in healthy adults. Cold Spring Harbor
Laboratory; 2017; Available from: \url{https://doi.org/10.1101/189886}

\hypertarget{ref-Salter_contamination_2014}{}
29. Salter SJ, Cox MJ, Turek EM, Calus ST, Cookson WO, Moffatt MF, et
al. Reagent and laboratory contamination can critically impact
sequence-based microbiome analyses. BMC Biology {[}Internet{]}. Springer
Nature; 2014;12. Available from:
\url{https://doi.org/10.1186/s12915-014-0087-z}

\hypertarget{ref-brim_microbiome_2013}{}
30. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
al. Microbiome analysis of stool samples from African Americans with
colon polyps. PloS One. 2013;8:e81352.
